Verifying Fine-Grained and Lock-Free Binary Search Trees

### 2021-12-14 concurrency/Verification of Concurrent Binary Search Tree.tex
Add content for delete in writeup

\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{iris}
%\usepackage{lstlangcoq}
\renewcommand{\lstlistingname}{Figure}

@@

>>>>

\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{iris}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

The code for the delete method is shown in ??. It takes the location of the root pointer
and a key as the arguments. A thread traverses the tree using hand-over-hand locking,
until the thread finds the key to be deleted. To remove the key, the tree is changed 
through an operation called pushdown left such that the node to be deleted points to an empty leaf
right child. To achieve that, the thread sets the right child's left child as the 
new right child of the to-be-deleted node and moves the right child's right child as 
the new parent of the to-be-deleted node. This operation is repeated until the right
child of the to-be-deleted node is an empty leaf node. Then, the thread will delete
the node and set the delete node's parent pointer to the left child of the deleted node.

The atomic specification for delete method can be written as follows:
$$\forall t.\ \langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(t,g)\rangle $$ 
$$\texttt{delete}(p,k)$$
$$\langle \nodeboxrep(p,g,g\_root)\ |\ \treerep(\texttt{delete}(t, k),g)\rangle $$

The delete method is unique compared to insert and lookup because the pushdown left 
operation can change the \textit{public} half range of more than one nodes. In addition,
the pushdown left operation is one of the main reasons why the ghost tree of the binary
search tree is not just a ``copy" of the pure tree: the functional model of pushdown\_left
is unable to account for the change of deleted node's parent's pointer from pointing
at the deleted node the to the deleted node's right child.

\usetikzlibrary{positioning}

\forestset{
   lab/.style={
        label={[anchor=west, scale=0.75] north east:#1},
    },
}
\begin{forest}
for tree={
    grow=south,
    circle, draw, minimum size=3ex, inner sep=1pt,
    s sep=3mm
        }
    % my label/.style={
        % label={[anchor=south west] north east:#1},
    % },
    [, phantom, for children={fit=band}, s sep+=20mm,
         before drawing tree={%
          tikz+={%
            \node (a) [inner sep=0pt, fit=(!1) (!1L) (!1F)] {};
            \node (b) [inner sep=0pt, fit=(!l) (!lL) (!lF)] {};
            \node [anchor=center] at ($(a.east)!1/2!(b.west)$) {$\longrightarrow\quad$};
          },
        },
        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
                [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
                    [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,40)$};}]
                    [55, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,75)$};}
                        [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,55)$};}]
                        [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
                    ]
                ]
                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
            ]
        ]
        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
                    [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(40,35)$};}]
                        [50, tikz={\node[right=0pt of .north east, scale=0.75]  {$(40,55)$};}]
                    ]
                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
                ]
                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
            ]
        ]
        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
                    [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
                        [40, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,50)$};}
                            [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
                            [,no edge, draw=none]
                        ]
                        [,no edge, draw=none]
                    ]
                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
                ]
                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
            ]
        ]
        [ 30, tikz={\node[right=0pt of .north east, scale=0.75]  {$(-\infty,\infty)$};}
            [25, tikz={\node[left=0pt of .north west, scale=0.75]  {$(-\infty,30)$};}]
            [75, tikz={\node[right=0pt of .north east, scale=0.75]  {$(30,\infty)$};}
                [55, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,75)$};}
                    [50, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,55)$};}
                        [35, tikz={\node[left=0pt of .north west, scale=0.75]  {$(30,35)$};}]
                        [,no edge, draw=none]
                    ]
                    [60, tikz={\node[right=0pt of .north east, scale=0.75]  {$(55,75)$};}]
                ]
                [80, tikz={\node[right=0pt of .north east, scale=0.75]  {$(75,\infty)$};}]
            ]
        ]
    ]
\end{forest}

\subsubsection{delete}

Similar with insert, delete starts by acquiring the root node lock, so that the thread
can access the information inside the lock invariant of the root node. Then, we prove
the while loop by showing that the precondition satisfies delete\_inv and the 
loop body preserves the loop invariant. The first two cases are similar to insert and
lookup in which the thread finds the target node, traversing by hand-over-hand locking.
Once the target node is located, the thread will call pushdown\_left and return after.
Using the specification of pushdown\_left, we show that the post-condition of 
pushdown\_left implies the loop invariant. Note that before going into pushdown\_left,
the thread is holding the lock to the target node and will return to the delete
method without holding any.

\subsubsection{pushdown\_left}

The proof to pushdown\_left is the key step of proving the delete method. 
First, we show that the precondition of pushdown\_left satisfies the precondition
of the loop invariant, which includes the assertion that the thread is holding the 
lock to the deleted node (i.e. the information inside the lock invariant is accessible)
and an atomic\_shift. Then, the thread accesses that information to acquire a second 
lock to the right child of the deleted node. Due to the hand-over-hand locking 
mechanism this will not cause a deadlock since the thread has already acquired the 
lock to the deleted node. Next, we prove that the action of deleting the node,
with sync\_commit by opening the global invariant and removing the relevant 
ghost names (the target node and its right sentinel node), satisfies the loop
invariant and the postcondition of pushdown\_left.

The second part of pushdown\_left is proving that the turn\_left operation satisfies
the loop invariant. Using a variant of the sync\_rollback rule, we show that the 
turn\_left operation preserves the BST, apart from changes in the ranges of two nodes,
the target node and its right child, and information about the nodes in those two nodes.
To satisfy the structure of the ranges, we only need to prove that the new range for
the target node is encompasses the old range of both nodes. This will imply that
the range of the target node's old parent includes the target node's right child 
new range as the right child becomes the new parent (and the old parent points to 
the right child).

One key step of the proof is proving that two ghost trees are equivalent if the union
of its ghost nodes are equal and the ghost tree is sorted. This allows the global invariant
to be preserved during the turn\_left operation in which no nodes were actually changed
except for the structure of the tree.
